Packing limits and developmental integration are not sufficient to explain stomatal anatomical evolution in flowering plants

### 2022-06-25 ms/ms.Rmd
r1

Functional and developmental constraints on phenotypic variation may cause traits to covary over millions of years and slow populations from reaching their adaptive optima. Alternatively, trait covariation may result from selective constraint if some trait combinations and generally maladaptive. Quantifying the relative contribution of functional, developmental, and selective constraints on phenotypic variation is a longstanding goal of macroevolution, but it is often difficult to distinguish different types of constraints. The stomatal anatomy of leaves with stomata on both surfaces (amphistomatous) present a unique opportunity to test the significance of functional and developmental constraints on phenotypyic evolution at broad phylogenetic scales. The key insight is that stomata on each leaf surface encounter the same functional and developmental constraints, but potentially very different selective constraints. If stomatal traits on each surface evolve differently, this implies that independent evolution is possible and functional or developmental constraints alone likely do not explain covariance between stomatal anatomical traits. Packing limits and cell-size-mediated developmental integration are hypothesized to constrain variation in stomatal anatomy. We tested this synthesizing data on stomatal density and length from amphistomatous leaves of `r n_tax3` terrestrial flowering plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from `r n_pair` phylogenetically independent contrasts using a robust Bayesian model. Contrary to packing limit and developmental integration hypotheses, stomatal anatomy on each surface diverged partially independently. To better understand the global variation in ecologically important traits like stomata will likely require adaptive explanations for what constrains phenotypic (co)variation.

@@

In this study, we will address packing problems and developmental integration, specific forms of functional and developmental constraint relevant to our study system, stomatal anatomy. We introduce these concepts generally in this paragraph. Packing a number objects into a finite space is common functional constraint on organisms. Regular geometries that appear in nature such as helices and hexagons (think DNA and honeycombs) are often optimal solutions to packing problems [@mackenzie_proving_1999; @maritan_optimal_2000]. Notice that functional constraint does not preclude selection, but the presence of a packing limit changes the range of possible phenotypes. Developmental integration is a form of developmental constraint on multivariate phenotypic evolution and we use them interchangeably in this study. Developmentally integrated traits have a "disposition for covariation" [@armbruster_integrated_2014], meaning that evolutionary divergence between lineages in one trait will be tightly associated with divergence in another trait. Allometry is a classic, albeit contested, example of developmental integration that may constrain phenotypic evolution [reviewed in @pelabon_evolution_2014]. Strong allometric covariation between traits within populations can constrain macroevolutionary divergence for long periods of time depending on the strength and direction of selection [@lande_quantitative_1979]. However, developmental integration does not necessarily hamper adaptation, and can even accelerate adaptive evolution when trait covariation is aligned with the direction of selection [@hansen_is_2003]. For example, fusion of floral parts increases their developmental integration which may increase the rate and precision of multivariate adaptation to specialist pollinators [Berg’s rule, @berg_general_1959; @berg_ecological_1960; @conner_raissa_2014; @armbruster_covariance_1999].

Biologists have studied phenotypic constraints for decades, but progress is challenging because many multivariate phenotypes are too complex or poorly understood to quantitatively distinguish functional, developmental, and selective constraints over macroevolutionary timescales. Stomatal anatomy on the leaves of land plants provide an exceptional opportunity because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of year of evolutionary history. A heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selective constraints may potentially differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selective constraints most likely explain at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological importance. The next sections provides background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important. Readers not interested in this background may skip ahead to hypotheses and predictions.

@@

Stomata are microscopic pores formed by a pair of guard cells that regulate gas exchange (CO$_2$ gain and water vapor loss) on the leaf or other photosynthetic surfaces of most land plants. Stomata originated once in the history of land plants around 500 Ma, diversified rapidly in density and size, and have been maintained in most lineages except some bryophytes and aquatic plants [recently reviewed @clark_origin_2022]. Stomata respond physiologically by opening and closing in response to light, humidity, temperature, circadian rhythm, and plant water status [@hetherington_role_2003; @lawson_guard_2020]. On this short time scale, the stomatal size, density, and distribution on a mature leaf do not change, so the maximum rate of is fixed. At somewhat longer timescales, the plant may respond to environmental cues such as light and CO$_2$ by altering stomatal anatomy in new leaves [@casson_influence_2008]. Physiological responses (aperture change) and plastic responses (new leaves with changed anatomy) may be alternative strategies for plants to acclimate to environmental change [@haworth_co-ordination_2013]. Finally, stomatal anatomy can evolve due to inherited changes in stomatal development. Plastic and genetic changes in stomatal anatomy are likely both ecologically important, but most studies do not use a common garden design that would tease apart their relative contribution.

@@

The adaptive significance of variation in stomatal ratio is uncertain, but we have some clues based on the distribution of hypo- and amphistomatous leaves. Despite the fact that amphistomy can increase photosynthesis, most leaves are hypostomatous, suggesting that the benefits of amphistomy. Amphistomy should increase photosynthesis most under saturating-light conditions where CO$_2$ supply limits photosynthesis. This may explain why amphistomatous leaves are most common in high light habitats [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017], especially in herbs [@muir_light_2018]. However, the light environment alone cannot explain why hypostomatous leaves predominate in shade plants [@muir_is_2019], suggesting that we need to understand the costs of upper stomata better. Upper stomata increase the susceptibility to rust pathogens in *Populus* [@mckown_association_2014; @mckown_role_2019; @fetter_growthdefense_2021]. Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits [@buckley_how_2015]. Other hypotheses about the adaptive significance of stomatal ratio are discussed in @muir_making_2015 and @drake_two_2019.

@@

Two salient features of stomatal anatomy have been recognized for decades but we do not yet understand the evolutionary forces that generate and maintain them. We denote these two features as “inverse size-density scaling” and “bimodal stomatal ratio” (Fig. \ref{fig:concepts}). Inverse size-density scaling refers to the negative interspecific correlation between the size of the stomatal apparatus and the density of stomata [@weiss_untersuchungen_1865; @franks_maximum_2009; @de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021]. Across species, leaves with smaller stomata tend to pack them more densely, but there is significant variation about this general trend (Fig. \ref{fig:concepts}a). Stomatal size and density determine the maximum stomatal conductance to CO$_2$ and water vapor but also take up space on the epidermis, which could be costly for both construction and maintenance. Natural selection should favor leaves that have enough stomata of sufficient size to supply CO$_2$ for photosynthesis. Hence leaves with few, small stomata and high photosynthetic rates do not exist because they would not supply enough CO$_2$. Conversely, excess stomata or extra large stomata beyond the optimum may result in stomatal interference where the CO$_2$ concentration gradient around one stomate merges with that of its neighbor [@zeiger_stomatal_1987; @lehmann_effects_2015], incur metabolic costs [@deans_optimization_2020], and/or risk hydraulic failure [@henry_stomatal_2019]. The distribution of stomata and density may therefore represent the combinations that ensure enough, but not too much, stomatal conductance. @franks_maximum_2009 further hypothesized that the evolution of small stomata in angiosperms enabled increased stomatal conductance while minimizing the epidermal area allocated to stomata.

A striking feature of the interspecific variation in stomatal ratio is that trait values are not uniformly distributed, but strongly bimodal \ref{fig:concepts}b. Bimodal stomatal ratio refers to the observation that the ratio of stomatal density on the adaxial (upper) surface to the density on the abaxial (lower) has distinct modes (Fig. \ref{fig:concepts}b). Amphistomy occurs most often in herbaceous plants from open, high light habitats [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017; @muir_light_2018]. @muir_making_2015 described bimodal stomatal ratio formally but the pattern is apparent in earlier comparative studies of the British flora (*cf.* @peat_comparative_1994 Fig 1).

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/concepts.pdf}
\caption{Two salient features of stomatal anatomy in land plants are the (a) inverse relationship between stomatal size and density and (b) the bimodal distribution of stomatal ratio. At broad phylogenetic scales, leaves with smaller stomata ($x$-axis, log-scale) tend to have greater stomatal density  ($x$-axis, log-scale), but there is a lot of variation about the overall trend indicated by the grey ellipse. Hypostomatous leaves (stomatal ratio = 0) are more common than amphistomatoues leaves, but within amphistomatous leaves, the density of stomata on each surface tends to be similar (stomatal ratio $\approx$ 0.5), which we refer to as bimodal stomatal ratio.}
\label{fig:concepts}
\end{figure}

@@

Given the significance of stomata for plant function and global vegetation modeling [@berry_stomata:_2010], we would like to understand what factors constrain their anatomical variation. Could packing limits and developmental integration explain? We limit our focus in this study to interspecific variation in mean trait values and do not study understand intraspecifc variation here.

Packing constraints and developmental integration could explain inverse size-density scaling. The density and size of a stomata on a planar leaf surface can be viewed as a packing problem where the total area allocated to stomata cannot exceed the total leaf area. This is a functional, or geometric, constraint because certain combinations of large size and high density are not physically possible. This functional constraint cannot explain why combinations of low density and small size are rare, but may explain why stomatal size must decrease when density increase as the leaf runs out of space. The packing limit of functional stomata is less than the entire leaf area, but the exact value is unclear. The realized upper limit is close to 1/3 or 1/2 [@de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021].

Guard cell size and spacing between stomata (the inverse of density) are developmentally intertwined because guard cells and epidermal pavement cells between stomata develop from the same meristem. Before guard cell meristemoids form via asymmetric cell division [@dow_patterning_2014], the size of guard and epidermal cells are influenced by meristematic cell volume and expansion. Evolutionary shifts in meristematic cell volume or expansion rate could cause both increased stomatal size and lower density because epidermal cells between stomata are larger [@brodribb_unified_2013]. For example, larger genomes increase meristematic cell volume [@simova_geometrical_2012], setting a lower bound on final cell volume. Although different expansion rates in guard and epidermal pavement cells can reduce the correlation in their final size, the fact that species with larger genomes tend toward having larger stomata and lower density may indicate an effect of development integration of stomatal anatomy [@beaulieu_genome_2008; @simonin_genome_2018; @roddy_scaling_2020]. Developmental integration in this case would not hinder adaptive evolution when the main axes of selection were aligned with the developmental correlation. For example, if higher maximum stomatal conductance were achieved primarily by increasing stomatal density and decreasing stomatal size as proposed by @franks_maximum_2009, then developmental integration might accelerate the response to selection compared to a case where stomatal size and density are completely independent.

@@

The overarching question is whether major features of stomatal anatomy in terrestrial angiosperms are consistent with packing constraints and developmental integration mediated by cell size. Since stomata on both surfaces of amphistomatous leaves are subject to the same functional and developmental constraints, if these constraints are most important we predict similar patterns of trait covariation on abaxial and adaxial surfaces. Conversely, if traits covary differently on each surface it would indicate that stomatal anatomical traits can evolve independently and selective constraints like contribute to covariation. Analogously, variation in the genetic correlation and interspecific divergence of sexually dimorphic traits in dioecious species demonstrate that integration is not fixed and can be modified by selection [@barrett_sexual_2013]. We framed specific hypotheses and predictions around how functional or developmental constraints might explain either inverse size-density scaling or bimodal stomatal ratio.

@@

If the developmental integration hypothesis is correct, it also implies stomatal size and density will diverge in concert on each surface because the developmental function is fixed. Therefore we predict that divergence of stomatal traits on one surface will be isometric with divergence in stomatal traits on the other surface. This type of developmental integration limits the expression of variation that could give rise to a bimodal stomatal ratio. Suppose that in the hypostomatous leaves, stomatal development is completely suppressed. In amphistomatous leaves, stomatal development proceeds identically on each surface because the developmental function is identical. This would lead to a tendency for equal density on each surface.

@@

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied [@bergmann_stomatal_2007; @dow_patterning_2014; @sack_developmental_2016]. We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We integrated and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to Notes A1. A graphical summary is provided in Fig. \@ref(fig:developmental-integration). We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division [@dow_patterning_2014]. The second parameter is stomatal index [@salisbury_i_1928; @sack_developmental_2016], which is determined by amplifying and spacing divisions after asymmetric cell division [@dow_patterning_2014]. When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density [see @lynch_genetics_1998 for key random variable algebra theorems].

@@

Data on stomatal anatomy are spread over a disparate literature and we have not attempted an exhaustive synthesis of amphistomatous leaf stomatal anatomy. We began our search by reviewing papers that cited key studies of amphistomy [@parkhurst_adaptive_1978; @mott_adaptive_1982; @muir_making_2015]. We supplemented these by searching Clarivate *Web of Science* for "guard cell length" because most studies that report guard cell length also report stomatal density, whereas the reverse is not true. We identified additional studies by reviewing the literature cited of papers we found and through opportunistic discovery. The final data set contained `r nrow(stomata_anatomy1)` observations of stomatal density and length from `r length(unique(stomata_anatomy1$source_id))` taxa and `r length(unique(stomata_anatomy1$source))` primary studies (Table \@ref(tab:sources)). However, many of these data were excluded if taxonomic name and phylogenetic placement could not be resolved (see below). Finally, we included some unpublished data. Stomatal size data were collected on grass species described in @pathare_increased_2020. We also included unpublished data on 14 amphistomatous wild tomato species (*Solanum* sect. *Lycopersicum* and sect. *Lycopersicoides*) grown in pots under outdoor summer Mediterranean conditions described in @muir_unpublished_2022. We took ab- and adaxial epidermal imprints using clear nail polish of the mid-portion of the lamina away from major veins on the terminal leaflet of the youngest, fully expanded leaf from 1-5 replicates per taxon. With a brightfield light microscope, we counted stomata in three 0.571 mm$^2$ fields of view and divided by the total area to estimate density. We measured the average guard cell length of 60 stomata, 20 per field of view, to estimate stomatal size. The data set is publicly available as an *R* package **ropenstomata** (https://github.com/cdmuir/ropenstomata). We collected data on genome size from the Angiosperm DNA C-values database [@leitch_angiosperm_2019; @pellicer_plant_2020]. When multiple ploidy levels were available for a taxon, we chose the lowest one for consistency. All data will be deposited on Dryad and archived on Zenodo upon publication.

@@

All hypotheses make predictions about trait (co)variance matrices or parameters derived from them (see Notes A1 and subsections below). Within and among species covariation is a hallmark of developmental integration [@armbruster_multilevel_1988], but other evolutionary processes also lead to covariance. Distinguishing between them requires deriving predictions and testing whether observed covariance is consistent with one hypothesis or another. We estimated the $4 \times 4$ covariance matrix of phylogenetically independent contrasts between log-transformed values of $\Delta \text{log}(D_\mathrm{ab})$, $\Delta \text{log}(D_\mathrm{ad})$, $\Delta \text{log}(L_\mathrm{ab})$, and $\Delta \text{log}(L_\mathrm{ad})$ using a distributional multiresponse robust Bayesian approach. See Table \ref{tab:traits} for variable definitions. We denote variances as $\text{Var}[\Delta \text{log(trait)}]$ and covariances as $\text{Cov}[\Delta \text{log(trait}_1),\Delta \text{log(trait}_2)]$. The statistical model also accounts for differences in mean trait values between surfaces. We used a multivariate $t$-distribution rather than a Normal distribution because estimates using the former are more robust to exceptional trait values [@lange_robust_1989]. We also estimated whether the variance in trait divergence increases with time. Under many trait evolution models (e.g. Brownian motion), interspecific variance increases through time. To account for this, we included time since taxon-pair divergence as an explanatory variable affecting the trait covariance matrix, but not the trait mean.

@@

We tested the whether the covariance between divergence in stomatal length and stomatal density on each leaf surface is the same. If size and density are developmentally integrated, we predict the covariance matrices will not be significantly different. Specifically, the 95\% HPD intervals of the difference in covariance parameters should not include 0 if:

@@

## Theory: from development integration to disintegration

>>>>

Functional and developmental constraints on phenotypic variation may cause traits to covary over millions of years and slow populations from reaching their adaptive optima. Alternatively, trait covariation may result from selective constraint if some trait combinations are generally maladaptive. Quantifying the relative contribution of functional, developmental, and selective constraints on phenotypic variation is a longstanding goal of macroevolution, but it is often difficult to distinguish different types of constraints. The stomatal anatomy of leaves with stomata on both surfaces (amphistomatous) present a unique opportunity to test the significance of functional and developmental constraints on phenotypyic evolution at broad phylogenetic scales. The key insight is that stomata on each leaf surface encounter the same functional and developmental constraints, but potentially very different selective constraints. If stomatal traits on each surface evolve differently, this implies that independent evolution is possible and functional or developmental constraints alone likely do not explain trait covariance. Packing limits and cell-size-mediated developmental integration are hypothesized to constrain variation in stomatal anatomy. We tested this by synthesizing data on stomatal density and length from amphistomatous leaves of `r n_tax3` terrestrial flowering plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from `r n_pair` phylogenetically independent contrasts using a robust Bayesian model. Contrary to packing limit and developmental integration hypotheses, stomatal anatomy on each surface diverged partially independently. To better understand the global variation in ecologically important traits like stomata will likely require adaptive explanations for what constrains phenotypic (co)variation.

@@

In this study, we will address packing problems and developmental integration, specific forms of functional and developmental constraint relevant to our study system, stomatal anatomy. We introduce these concepts generally in this paragraph. Packing a number objects into a finite space is a common functional constraint on organisms. Regular geometries that appear in nature such as helices and hexagons (think DNA and honeycombs) are often optimal solutions to packing problems [@mackenzie_proving_1999; @maritan_optimal_2000]. Notice that functional constraint does not preclude selection, but the presence of a packing limit changes the range of possible phenotypes. Developmental integration is a form of developmental constraint on multivariate phenotypic evolution and we use them interchangeably in this study. Developmentally integrated traits have a "disposition for covariation" [@armbruster_integrated_2014], meaning that evolutionary divergence between lineages in one trait will be tightly associated with divergence in another trait. Allometry is a classic, albeit contested, example of developmental integration that may constrain phenotypic evolution [reviewed in @pelabon_evolution_2014]. Strong allometric covariation between traits within populations can constrain macroevolutionary divergence for long periods of time depending on the strength and direction of selection [@lande_quantitative_1979]. However, developmental integration does not necessarily hamper adaptation, and can even accelerate adaptive evolution when trait covariation is aligned with the direction of selection [@hansen_is_2003]. For example, fusion of floral parts increases their developmental integration which may increase the rate and precision of multivariate adaptation to specialist pollinators [Berg’s rule, @berg_general_1959; @berg_ecological_1960; @conner_raissa_2014; @armbruster_covariance_1999].

Biologists have studied phenotypic constraints for decades, but progress is challenging because many multivariate phenotypes are too complex or poorly understood to quantitatively distinguish functional, developmental, and selective constraints over macroevolutionary timescales. Stomatal anatomy on the leaves of flowering plants provide an exceptional opportunity because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of years of evolutionary history. For this purpose, a heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selective constraints may differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selective constraints most likely explain at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological significance. The next sections provide background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important. Readers not interested in this background may skip ahead to hypotheses and predictions.

@@

Stomata are microscopic pores formed by a pair of guard cells that regulate gas exchange (CO$_2$ gain and water vapor loss) on the leaf or other photosynthetic surfaces of most land plants. Stomata originated once in the history of land plants around 500 Ma, diversified rapidly in density and size, and have been maintained in most lineages except some bryophytes and aquatic plants [recently reviewed in @clark_origin_2022]. Stomata respond physiologically by opening and closing in response to light, humidity, temperature, circadian rhythm, and plant water status [@hetherington_role_2003; @lawson_guard_2020]. On this short time scale, the stomatal size, density, and distribution on a mature leaf do not change, so the maximum rate of is fixed. At somewhat longer timescales, the plant may respond to environmental cues such as light and CO$_2$ by altering stomatal anatomy in new leaves [@casson_influence_2008]. Physiological responses (aperture change) and plastic responses (new leaves with changed anatomy) may be alternative strategies for plants to acclimate to environmental change [@haworth_co-ordination_2013]. Finally, stomatal anatomy can evolve due to inherited changes in stomatal development. Plastic and genetic changes in stomatal anatomy are both ecologically important, but most studies do not use a common garden design that would tease apart their relative contribution.

@@

The adaptive significance of variation in stomatal ratio is uncertain, but we have some clues based on the distribution of hypo- and amphistomatous leaves. Despite the fact that amphistomy can increase photosynthesis, most leaves are hypostomatous. Amphistomy should increase photosynthesis most under saturating-light conditions where CO$_2$ supply limits photosynthesis. This may explain why amphistomatous leaves are most common in high light habitats [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017], especially in herbs [@muir_light_2018]. However, the light environment alone cannot explain why hypostomatous leaves predominate in shade plants [@muir_is_2019], suggesting that we need to understand the costs of upper stomata better. Upper stomata increase the susceptibility to rust pathogens in *Populus* [@mckown_association_2014; @mckown_role_2019; @fetter_growthdefense_2021]. Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits [@buckley_how_2015]. Other hypotheses about the adaptive significance of stomatal ratio are discussed in @muir_making_2015 and @drake_two_2019.

@@

Two major features of stomatal anatomy have been recognized for decades but we do not yet understand the evolutionary forces that generate and maintain them. We denote these two features as “inverse size-density scaling” and “bimodal stomatal ratio” (Fig. \ref{fig:concepts}). Inverse size-density scaling refers to the negative interspecific correlation between the size of the stomatal apparatus and the density of stomata [@weiss_untersuchungen_1865; @franks_maximum_2009; @de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021]. Across species, leaves with smaller stomata tend to pack them more densely, but there is significant variation about this general trend (Fig. \ref{fig:concepts}a). Stomatal size and density determine the maximum stomatal conductance to CO$_2$ and water vapor but also take up space on the epidermis, which could be costly for both construction and maintenance. Natural selection should favor leaves that have enough stomata of sufficient size to supply CO$_2$ for photosynthesis. Hence leaves with few, small stomata and high photosynthetic rates do not exist because they would not supply enough CO$_2$. Conversely, excess stomata or extra large stomata beyond the optimum may result in stomatal interference where the CO$_2$ concentration gradient around one stomate merges with that of its neighbor [@zeiger_stomatal_1987; @lehmann_effects_2015], incur metabolic costs [@deans_optimization_2020], and/or risk hydraulic failure [@henry_stomatal_2019]. The distribution of stomatal size and density may therefore represent the combinations that ensure enough, but not too much, stomatal conductance. @franks_maximum_2009 further hypothesized that the evolution of small stomata in angiosperms enabled increased stomatal conductance while minimizing the epidermal area allocated to stomata.

A striking feature of the interspecific variation in stomatal ratio is that trait values are not uniformly distributed, but strongly bimodal (Fig.  \ref{fig:concepts}b). Bimodal stomatal ratio refers to the observation that the ratio of stomatal density on the adaxial (upper) surface to the density on the abaxial (lower) has distinct modes (Fig. \ref{fig:concepts}b). Amphistomy occurs most often in herbaceous plants from open, high light habitats [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017; @muir_light_2018]. @muir_making_2015 described bimodal stomatal ratio formally but the pattern is apparent in earlier comparative studies of the British flora [*cf.* @peat_comparative_1994 Fig. 1].

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/concepts.pdf}
\caption{Two salient features of stomatal anatomy in flowering plants are the (a) inverse relationship between stomatal size and density and (b) the bimodal distribution of stomatal ratio. At broad phylogenetic scales, leaves with smaller stomata ($x$-axis, log-scale) tend to have greater stomatal density  ($x$-axis, log-scale), but there is a lot of variation about the overall trend indicated by the grey ellipse. Hypostomatous leaves (stomatal ratio = 0) are more common than amphistomatoues leaves, but within amphistomatous leaves, the density of stomata on each surface tends to be similar (stomatal ratio $\approx$ 0.5), which we refer to as bimodal stomatal ratio.}
\label{fig:concepts}
\end{figure}

@@

Given the significance of stomata for plant function and global vegetation modeling [@berry_stomata:_2010], we would like to understand what factors constrain their anatomical variation. Here we focus on interspecific variation in mean trait values rather than intraspecifc variation. Packing constraints and developmental integration could explain inverse size-density scaling. The density and size of a stomata on a planar leaf surface can be viewed as a packing problem where the total area allocated to stomata cannot exceed the total leaf area. This is a functional, or geometric, constraint because certain combinations of large size and high density are not physically possible. This functional constraint cannot explain why combinations of low density and small size are rare, but may explain why stomatal size must decrease when density increases as the leaf runs out of space. The packing limit of functional stomata is less than the entire leaf area, but the exact value is unclear. The realized upper limit is close to 1/3 or 1/2 [@de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021] for the species' mean, not an individual leaf.

Guard cell size and spacing between stomata (the inverse of density) are developmentally intertwined because guard cells and epidermal pavement cells between stomata develop from the same meristem. Before guard cell meristemoids form via asymmetric cell division [@dow_patterning_2014], the size of guard and epidermal cells are influenced by meristematic cell volume and expansion. Evolutionary shifts in meristematic cell volume or expansion rate could cause both increased stomatal size and lower density because epidermal cells between stomata are larger [@brodribb_unified_2013]. For example, larger genomes increase meristematic cell volume [@simova_geometrical_2012], setting a lower bound on final cell volume. Although different expansion rates in guard and epidermal pavement cells can reduce the correlation in their final size, the fact that species with larger genomes tend toward having larger stomata and lower density may indicate an effect of development integration on stomatal anatomy [@beaulieu_genome_2008; @simonin_genome_2018; @roddy_scaling_2020]. Developmental integration in this case would not necessarily hinder adaptive evolution if the main axes of selection were aligned with the developmental correlation. For example, if higher maximum stomatal conductance were achieved primarily by increasing stomatal density and decreasing stomatal size as proposed by @franks_maximum_2009, then developmental integration might accelerate the response to selection compared to a case where stomatal size and density are completely independent.

@@

The overarching question is whether major features of stomatal anatomy in terrestrial angiosperms are consistent with packing constraints and/or developmental integration mediated by cell size. Since stomata on both surfaces of amphistomatous leaves are subject to the same functional and developmental constraints, if these constraints are most important we predict similar patterns of trait covariation on abaxial and adaxial surfaces. Conversely, if traits covary differently on each surface it would indicate that stomatal anatomical traits can evolve independently and selective constraints likely contribute to covariation. Analogously, variation in the genetic correlation and interspecific divergence of sexually dimorphic traits in dioecious species demonstrate that integration is not fixed and can be modified by selection [@barrett_sexual_2013]. We framed specific hypotheses and predictions around how functional or developmental constraints might explain either inverse size-density scaling or bimodal stomatal ratio.

@@

If the developmental integration hypothesis is correct, it also implies stomatal size and density will diverge in concert on each surface because the developmental function is fixed. Therefore we predict that divergence of stomatal traits on one surface will be isometric with divergence in stomatal traits on the other surface. This type of developmental integration limits the expression of variation and could give rise to a bimodal stomatal ratio. Suppose that in hypostomatous leaves, stomatal development is completely suppressed. In amphistomatous leaves, stomatal development proceeds identically on each surface because the developmental function is identical. This would lead to a tendency for equal density on each surface.

@@

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied [@bergmann_stomatal_2007; @dow_patterning_2014; @sack_developmental_2016]. We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We combined and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to Notes A1. A graphical summary is provided in Fig. \@ref(fig:developmental-integration). We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division [@dow_patterning_2014]. The second parameter is stomatal index [@salisbury_i_1928; @sack_developmental_2016], which is determined by amplifying and spacing divisions after asymmetric cell division [@dow_patterning_2014]. When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density [see @lynch_genetics_1998 for key random variable algebra theorems].

@@

Data on stomatal anatomy are spread over a disparate literature and we have not attempted an exhaustive synthesis of amphistomatous leaf stomatal anatomy. We began our search by reviewing papers that cited key studies of amphistomy [@parkhurst_adaptive_1978; @mott_adaptive_1982; @muir_making_2015]. We supplemented these by searching Clarivate *Web of Science* for "guard cell length" because most studies that report guard cell length also report stomatal density, whereas the reverse is not true. We identified additional studies by reviewing the literature cited of papers we found and through opportunistic discovery. The final data set contained `r nrow(stomata_anatomy1)` observations of stomatal density and length from `r length(unique(stomata_anatomy1$source_id))` taxa and `r length(unique(stomata_anatomy1$source))` primary studies (Table \@ref(tab:sources)). However, many of these data were excluded if taxonomic name and phylogenetic placement could not be resolved (see below). Finally, we included some unpublished data. Stomatal size data were collected on grass species described in @pathare_increased_2020. We also included unpublished data on 14 amphistomatous wild tomato species (*Solanum* sect. *Lycopersicum* and sect. *Lycopersicoides*) grown in pots under outdoor summer Mediterranean conditions as described in @muir_unpublished_2022. We took ab- and adaxial epidermal imprints using clear nail polish of the mid-portion of the lamina away from major veins on the terminal leaflet of the youngest, fully expanded leaf from 1-5 replicates per taxon. With a brightfield light microscope, we counted stomata in three 0.571 mm$^2$ fields of view and divided by the total area to estimate density. We measured the average guard cell length of 60 stomata, 20 per field of view, to estimate stomatal size. The data set is publicly available as an *R* package **ropenstomata** (https://github.com/cdmuir/ropenstomata). We collected data on genome size from the Angiosperm DNA C-values database [@leitch_angiosperm_2019; @pellicer_plant_2020]. When multiple ploidy levels were available for a taxon, we chose the lowest one for consistency. All data will be deposited on Dryad and archived on Zenodo upon publication.

@@

All hypotheses make predictions about trait (co)variance matrices or parameters derived from them (see Notes A1 and subsections below). Within and among species covariation is a hallmark of developmental integration [@armbruster_multilevel_1988], but other evolutionary processes also lead to covariance. Distinguishing between them requires deriving predictions and testing whether observed covariance is consistent with one hypothesis or another. We estimated the $4 \times 4$ covariance matrix of phylogenetically independent contrasts between log-transformed values of $\Delta \text{log}(D_\mathrm{ab})$, $\Delta \text{log}(D_\mathrm{ad})$, $\Delta \text{log}(L_\mathrm{ab})$, and $\Delta \text{log}(L_\mathrm{ad})$ using a distributional multiresponse robust Bayesian approach. See Table \ref{tab:traits} for variable definitions. We denote variances as $\text{Var}[\Delta \text{log(trait)}]$ and covariances as $\text{Cov}[\Delta \text{log(trait}_1),\Delta \text{log(trait}_2)]$. We used a multivariate $t$-distribution rather than a Normal distribution because estimates using the former are more robust to exceptional trait values [@lange_robust_1989]. We also estimated whether the variance in trait divergence increases with time. Under many trait evolution models (e.g. Brownian motion), interspecific variance increases through time. To account for this, we included time since taxon-pair divergence as an explanatory variable affecting the trait covariance matrix.

@@

We tested whether the covariance between divergence in stomatal length and stomatal density on each leaf surface is the same. If size and density are developmentally integrated, we predict the covariance matrices will not be significantly different. Specifically, the 95\% HPD intervals of the difference in covariance parameters should not include 0 if:

@@

## Theory: from developmental integration to disintegration
